The reason intuitionistic logic was the focus is that it lends itself
very nicely to being interpreted as a system of computation.  That's
the entire point behind the BHK-interpretation and the work of
Brouwer.  This, it seemed, was not the case for classical logic, until
Michel Parigot constructed the $\lambda\mu$-calculus in 1992
\cite{Parigot:1992}.  Parigot was able to define a classical sequent
calculus called free deduction which had a cut-elimination procedure
validating the cut-theorem for classical logic \cite{Parigot:1992b}.
This allowed for Parigot to define a computational perspective of free
deduction which he called the $\lambda\mu$-calculus.  We now briefly
introduce the $\lambda\mu$-calculus.  The syntax and reduction rules
are in Fig.~\ref{fig:lamu_syntax}.

  \begin{figure}
  \begin{center}
    \begin{tabular}{lll}
      Syntax: 
      \vspace{10px} \\
      \begin{math}
        \begin{array}{lll}
          [[T]],[[A]],[[B]],[[C]] & ::= & [[X]]\,|\,[[_|_]]\,|\,[[A -> B]]\\
          [[t]] & ::= & [[x]]\,|\,[[\x.t]]\,|\,[[\m a.s]]\,|\,[[t1 t2]]\\
          [[s]] & ::= & [[ [a]t]] 
        \end{array}
      \end{math} \\
      & \\
      Full $\beta$-reduction: \\
      \begin{mathpar}
        \LamudruleRXXBeta{}       \and
        \LamudruleRXXStruct{}     \and
        \LamudruleRXXRenaming{}   \and
        \LamudruleRXXLam{}        \and
        \LamudruleRXXMu{}         \and
        \LamudruleRXXNaming{}     \and
        \LamudruleRXXAppOne{}     \and
        \LamudruleRXXAppTwo{}
      \end{mathpar}
    \end{tabular}
  \end{center}

  \caption{Syntax and reduction rules for the $\lambda\mu$-calculus}
  \label{fig:lamu_syntax}
\end{figure}

We can think of the language of the $\lambda\mu$-calculus as an
extension of the $\lambda$-calculus.  We extend it with two new
operators.  The first is the $\mu$-abstraction $[[\m a.s]]$ where
$[[a]]$ is called a co-variable, an output port, or an output
variable.  We call the $\mu$-abstraction a control operator.  This
name conveys the fact that the $\mu$-abstraction has the ability to
control whether a value is returned or placed into its bound output
port.  The body of the $\mu$-abstraction must be a term called a
statement denoted by the metavariable $[[s]]$.  Statements have the
form $[[ [a]t]]$.  We can think of this as assigning (or naming) an
output port to a term.  Now we extend the reduction rules with two new
reduction rules and two new congruence rules for the $\mu$-abstraction
and naming operator. The \Lamudrulename{R\_Struct} rule is called the
structural reduction rule.  This allows one to target reduction to a
named subterm of the body of the $\mu$-abstraction.  This rule uses a
special substitution operation $[[ [t /* a]s]]$ which says to replace
every subterm of $[[s]]$ matching the pattern $[[ [a]t']]$ with 
$[[ [a](t' t)]]$.  We may also write $[[ [t /* a]t']]$ for the similar
operation on terms. This is called structural substitution.

As we said above the language of the $\lambda\mu$-calculus is an
extension of the $\lambda$-calculus, but its type assignment is very
different than STLC.  The type assignment rules are defined in
Fig.~\ref{fig:lamu_typing}.
\begin{figure}
  \begin{center}
    \begin{mathpar}
      \LamudruleVar{} \and
      \LamudruleLam{} \and
      \LamudruleMu{}  \and
      \LamudruleApp{} \and
      \LamudruleNameApp{}
    \end{mathpar}
  \end{center}
  \caption{Type-checking algorithm for the $\lambda\mu$-calculus}
  \label{fig:lamu_typing}
\end{figure}
Right away we can see a difference in the form of judgment.  We now
have $[[e : G |- D]]$ rather than $\Gamma \vdash t :
\tau$.  The former is in sequent form. This is the orignal presentation used
by Parigot. The feature of this is that it make it easy to see when the set of assumptions and 
conclusions are modified \footnote{This is not the only formalization we could have
used.  See \cite{Curien:2000} for another example which is closer to the style
we have been using for the earlier type theories.} Think of $[[e : G |- D]]$ as $[[e]]$ being a
witness\footnote{Actually, ``the witness'', because typing in unique.}
of the sequent $[[G]] [[|-]] [[D]]$.  Just as in the other type
theories we have seen, $[[G]]$ is the typing context or the set of
assumptions (input ports).  Keeping to the style of Parigot we denote elements of $[[G]]$ by $[[A]]^{[[x]]}$ instead
of $[[x]] : [[A]]$.  The environment $[[D]]$ is either empty
$[[.]]$, a formula $[[A]]$, one or more co-assumptions or output
ports, or a formula $[[A]]$ followed by one or more output ports.  Negation is defined
in the same way as it is in intuitionistic logic.  That is $[[{- A}]] =^{def} [[A -> _|_]]$.
Note that in $[[D]]$ we always have $[[_|_ a]]$ (false) and in $[[G]]$ we
always have $\top^[[x]]$ (true) where $\top \equiv [[_|_ -> _|_]]$ for any
$[[D]]$ and $[[G]]$ trivially.  We often leave these left implicit to
make the presentation clean unless absolutely necessary.  These two
facts hold because a sequent $[[A1]]^{[[x1]]},\cdots,[[Ai]]^{[[xi]]}
[[|-]] [[B]],[[B1]]^{[[a1]]},\cdots,[[Bi]]^{[[ai]]}$ 
can be interpreted as 
$([[A1]]^{[[x1]]} \land \cdots \land [[Ai]]^{[[xi]]}) \implies ([[B]] \lor [[B1]]^{[[a1]]} \lor \cdots \lor [[Bi]]^{[[ai]]})$
where $\implies$ is implication.  Using this interpretation we can see that
adding true to the left and/or false to the right does not impact the logical
truth of the statement.  This implies the following lemma.
\begin{lemma}
  \label{lemma:lamu_neg}
  The following rules are admissible w.r.t. the $\lambda\mu$-calculus:
  \begin{center}
    \begin{math}
      \begin{array}{lll}
        $$\mprset{flushleft}
        \inferrule* [right=BtmInt] {
          [[a fresh in D]]
          \\
            [[s : G |- D]]
        }{[[\m a.s : G |- _|_, D]]}
        &
        $$\mprset{flushleft}
        \inferrule* [right=BtmElim] {
          [[a fresh in D]]
          \\
            [[t : G |- _|_,D]]          
        }{[[ [a]t : G |- D]]}
      \end{array}
    \end{math}
  \end{center}
\end{lemma}

The $\lambda\mu$-calculus is a classical type theory so it should be
the case that the law of excluded middle (LEM), $[[A]] \lor [[{-A}]]$,
holds, or equivalently the law of double negation (LDN) $[[{- {- A}} -> A]]$.
Since we do not have disjunction as a primitive we show LDN.
Before showing the derivation of the LDN we first define some
derived rules for handling negation and sequent manipulation rules.
The following definition defines all derivable rules.  We will take
these as primitive to make things cleaner.  We do not show the
derivations here, because they are rather straightforward.
\begin{lemma}
  \label{lemma:lamu_derived_rules}
  The following rules are derivable using the typing rules and the rules of
  Lemma~\ref{lemma:lamu_neg}:
  \begin{center}
    \begin{math}
      \begin{array}{lll}
        $$\mprset{flushleft}
        \inferrule* [right=NegInt1] {
          [[t : G,A x |- _|_,D]]
        }{[[\x.t:G |- {-A},D]]}
        &
        $$\mprset{flushleft}
        \inferrule* [right=NegElim1] {
          [[t1 : G |- {-A},D]]
          \\
          [[t2 : G |- A, D]]
        }{[[t1 t2 : G |- _|_,D]]}\\       
        &\\
        $$\mprset{flushleft}
        \inferrule* [right=NegInt2] {
          [[a fresh in D]]
          \\
          [[s : G, A x |- D]]
        }{[[\x.\m a.s : G |- {-A},D]]}
        &
        $$\mprset{flushleft}
        \inferrule* [right=NegElim2] {
          [[t1 : G |- {-A},D]]
          \\
          [[t2 : G |- A, D]]
        }{[[t1 t2 : G |- D]]}
      \end{array}
    \end{math}
  \end{center}
\end{lemma}
We are now in the state where we can prove $[[{-{-A}} -> A]]$ in the $\lambda\mu$-calculus.
\begin{example}
  \label{ex:lamu_ldn}
  In this example we prove $[[{-{-A}} -> A]]$.  Suppose $D$ is the
  following derivation:
  \begin{center}
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right=\Lamudrulename{NegElim1}] {
        $$\mprset{flushleft}
        \inferrule* [right=\Lamudrulename{Var}] {
          \ 
        }{[[y : {-{-A}} y |- {-{-A}}, A a]]}
        \\
        $$\mprset{flushleft}
        \inferrule* [right=\Lamudrulename{NegInt2}] {
          $$\mprset{flushleft}
          \inferrule* [right=\Lamudrulename{NameApp}] {
            $$\mprset{flushleft}
            \inferrule* [right=\Lamudrulename{Var}] {
              \ 
            }{[[x : {-{-A}} y, A x |- A]]}
          }{[[ [a]x : {-{-A}} y, A x |- A a]]}
        }{[[\x.\m b.[a]x : {-{-A}} y |- {-A}, A a]]}
      }{[[y (\x.\m b.[a]x) : {-{-A}} y |- _|_,A a]]}
    \end{math}
  \end{center}
  Then the final proof is as follows:
  \begin{center}    
    \begin{math}
      $$\mprset{flushleft}
      \inferrule* [right=\Lamudrulename{Lam}] {
        $$\mprset{flushleft}
        \inferrule* [right=\Lamudrulename{Mu}] {
          $$\mprset{flushleft}
          \inferrule* [right=\Lamudrulename{BtmElim}] {
            D
          }{[[ [b'](y (\x.\m b.[a]x)) : {-{-A}} y |- A a]]}
        }{[[\m a.[b'](y (\x.\m b.[a]x)):{-{-A}} y |- A]]}
      }{[[\y.\m a.[b'](y (\x.\m b.[a]x)):. |- {-{-A}} -> A]]}
    \end{math}
  \end{center}
\end{example}
\noindent
In the above example we leave out freshness constraints to make the
presentation cleaner.  This example shows that the
$\lambda\mu$-calculus really is classical.  So from the logical
perspective of computation we gain classical reasoning, but do we
gain anything programmatically?  It turns out that we do.  We can
think of the $\mu$-abstraction and naming application as continuations
which allow us to define exceptions.  In fact a great way of thinking
about the $\mu$-abstraction $[[\m a.[b]t]]$ is due to Geuvers et al.:
\begin{center}
  \begin{quote}
    \emph{From a computational point of view one should think of $[[\m a.[b]t]]$
    as a combined operation that catches exceptions labeled $[[a]]$
    in $[[t]]$ and throws the results of $[[t]]$ to $[[b]]$.} \cite{Geuvers:2012}
  \end{quote}
\end{center}
Using this point of view we can define $[[catch a t]]$ and $[[throw a t]]$.
\begin{definition}
  \label{def:lamu_catch_throw}
  The following defines exceptions within the $\lambda\mu$-calculus:
  \begin{center}
    \begin{math}
      \begin{array}{lll}
        [[catch a t]] & := & [[\m a.[a]t]]\\
        & \\
        [[throw a t]] & := & [[\m b.[a]t]] \text{, where } [[b]] \text{ is fresh }
      \end{array}
    \end{math}
  \end{center}
\end{definition}
Using our reduction rules with the addition of $[[\m a.[a]t ~> t]]$
provided that $[[a]]$ is fresh in $[[t]]$\footnote{This is sometimes called
$\eta$-reduction for control operators.},  we can easily define some
nice reduction rules for these definitions. 
\begin{definition}
  \label{def:lamu_catch_throw_red}
  Reduction rules for exceptions:
  \begin{center}
    \begin{math}
      \begin{array}{lll}
        [[catch a (throw a t) ~> catch a t]]\\
        & \\
        [[throw a (catch b t) ~> throw a ([a / b]t)]]
      \end{array}
    \end{math}
  \end{center}  
\end{definition}
\noindent
There are other reductions one might want.  For the others and an
extension of the $\lambda\mu$-calculus see \cite{Geuvers:2012}.
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: ../thesis.tex
%%% End: 
